---
title: "Ch5 Statistical Power"
author: "Lijia Wang"
date: "2020-07-29"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

This document is a summary of Chapter 5, "Statstical Power" in Neale et. al.. We will be recreating some of the examples given in the book chapter. Here I will try to explain and recreate table 5.5 on "Power to detect the effect of disrruptive selection on HWE in $n = 500$ and $\alpha = 0.05$"
```{r, echo=FALSE}
library(knitr)
```

### Setting up the HWE
A quick refresher on Hardy-Weinberg equilibrium: 
$$ p + q = 1 $$
$$ p^2 + 2pq + q^2 = 1 $$

We can use likelihood-ratio testing to detect violation of Hardy-Weinberg equilibrium given disruptive selection. We will first model it with a multinomial distribution Multi($\textbf{X}:n, \theta$): 

$$ \frac{n!}{x_1! x_2! ... x_k!} \theta^{x_1}_{1} \theta^{x_2}_{2} ... \theta^{x_k}_{k} $$
where k is the number of outcomes generated with fixed probabilities $\theta_1, \theta_2, ... \theta_k$. 

In a random sample of size $n = \Sigma x_i$, the genotype frequencies for AA, Aa, and aa can be written as:
$$ \textit{f}_1 = \frac{p^2}{t} $$
$$ \textit{f}_2 = \frac{2pq-r}{t} $$
$$ \textit{f}_3 = \frac{q^2}{t} $$

$$ t = p^2 + (2pq-r) + q^2 $$

Note the "-r" term that we typically do not see in the classical H-W equations. This is the non-centrality parameter (NCP) which we will use to model disruptions with. If $r=0$, then $\lambda=0$ and we are at H-W equilibrium


### Generating ML equations

To generate the ML equation, we can simply take the log of the likelihood equation of the multinomial distribution (as described above), and adapt it to the 3 genotypes AA, Aa, and aa, thus $k=3$. The LL equation would look like: 

$$LL_A = -2(log \frac{n!}{x_1! x_2! x_3!} + log(\theta_{A_1})^{x_1} + log(\theta_{A_2})^{x_2} + log(\theta_{A_3})^{x_3})$$
$$= -2(c + {x_1}log(\theta_{A_1}) + {x_2}log(\theta_{A_2}) + {x_3}log(\theta_{A_3})$$
Under the alternative hypothesis, only one among the three $\theta$ values is constrained (e.g. $\theta_{A3} = 1-\theta_{A1} - \theta_{A2}$). Under the more parsimonious null hypothesis, the frequencies must obey the HWE, therefore satisfying: 

$$ p_0 = \frac{2x_1 + x_2}{2n} $$
$$ q_0 = 1-p_0 $$
$$\theta_{01} = p_0^2 $$
$$\theta_{02} = 2p_0 q_0 $$
$$\theta_{03} = q_0^2 $$

But the ML equation is almost the exact same: 

$$LL_0= -2(c + {x_1}log(\theta_{0_1}) + {x_2}log(\theta_{0_2}) + {x_3}log(\theta_{0_3})$$

Lastly, the test statistic is calculated by: 
$$ T = LL_0(\theta_{0}; \textbf{X},n) - LL_A(\theta_{A}; \textbf{X},n) $$
And distributed with \chi^2(1,\lambda). 

### Analyzing statistical power
```{r}
# Code adopted from Neale 2008, Panel 5.2
N = 500
p = c(0.1, 0.3, 0.5)
r = c(0.01, 0.05, 0.1)    # non-centrality parameter
q = 1-p

mat = matrix(ncol = 4, nrow = 9)
row = 1

for (indp in 1:3) {
  for (indr in 1:3) {
    pv = p[indp]
    rv = r[indr]
    qv = q[indp]
    
    # model population distribution and distortion
    fr = c(pv^2, 2*pv*qv-rv, qv^2)
    fr = fr/sum(fr)                   # normalization 
    onfr = round(N*fr)                # observed frequency
    
    # estimate p and q to expected frequency under H-W
    pe = (onfr[1]*2 + onfr[2])/(2*N)
    qe = 1-pe
    efr = c(pe^2, 2*pe*qe, qe^2)
    logLA = 0
    for (i in 1:3) {
      logLA = logLA + onfr[i]*log(fr[i])
    }
    logL0 = 0
    for (i in 1:3) {
      logL0 = logL0 + onfr[i]*log(efr[i])
    }
    
    logLA = -2*logLA
    logL0 = -2*logL0
    lambda = logL0 - logLA                  # test statistic is chisq(1, lambda)
    alpha = 0.5
    df = 1
    c = qchisq(alpha, df, lower.tail = F)                     # critical value
    power = pchisq(c, df = df, ncp = lambda, lower.tail = F)  # 1-beta, power
    beta = 1 - power                                          # beta
  
    mat[row,] = c(p[indp], r[indr], lambda, power)
    row = row + 1
  }
  colnames(mat) = c('Frequency p','Effect size r', 'lambda', 'power')
}
kable(mat)
```


```{r workflow commands, eval=FALSE, echo=FALSE}
wflow_status()
wflow_publish(c('analysis/index.Rmd', 'analysis/statistical_power.Rmd'), 'ch5 table 5.5 recreation + HWE example explanation')
wflow_publish('analysis/statistical_power.Rmd', 'fixing up some latex code')
wflow_git_push()
```
